Efficient implementation of penalized models and their application to genetic risk prediction

### 2018-06-12 paper/draft.tex
minor text changes

In this paper, we present an efficient implementation of such penalized joint estimation, allowing for practical application of penalized logistic regression on modern datasets including hundreds of thousands of individuals. Our implementation of penalized logistic regression also includes nearly-optimal choices for hyper-parameters. The choice of hyper-parameters for a method is very important as it can dramatically impact the predictive performance of the method (e.g. the p-value threshold in C+T).

We compare the performance of penalized logistic regression to the C+T method and to a derivation of random forests. The penalized logistic regression consistently achieves higher predictive performance than the two other methods while being faster. We also found that improvement in predictive performance is more pronounced when there are few effects located in nearby genomic regions with correlated SNPs; AUC values increase from 83\% with C+T to 92.5\% with penalized logistic regression. We confirmed these results in a data analysis of a case-control study for celiac disease where penalized logistic regression and the standard C+T method achieve AUC of 89\% and of 82\%.

@@

Many methods have been developed to predict disease status, or more generally any phenotype, based on SNP information. A commonly used method, called ``P+T'' or ``C+T'' (which stands for ``Clumping and Thresholding'') -- or even just PRS -- is used to derive PRSs from results of Genome-Wide Association Studies (GWASs) \cite[]{chatterjee2013projecting,dudbridge2013power,evans2009harnessing,purcell2009common,wray2007prediction}. This technique uses GWAS summary statistics only, allowing for trivial and fast implementation. However, the approach also has limitations. In particular, previous studies showed that predictive performance of the C+T method is very sensitive to the threshold of inclusion of SNPs, depending on the disease architecture \cite{ware2017heterogeneity}.
Linear Mixed-Models (LMMs) are another widely-used method in fields such as plant and animal breeding or for predicting highly heritable quantitative human phenotypes such as height \cite[]{lello2017accurate,yang2010common}. Yet, models resulting from LMM, known e.g.\ as ``gBLUP'', are not optimal for predicting disease status based on genotypes \cite[]{abraham2013performance}. Moreover, these methods and their derivatives are often computationally demanding, both in terms of memory and time required, which makes them unlikely to be used for prediction on very large datasets \cite[]{golan2014effective,maier2015joint,speed2014multiblup,zhou2013polygenic}.
Finally, statistical learning methods have also been used to derive PRSs for complex human diseases by jointly estimating SNP effects. Such methods include joint logistic regression, Support Vector Machine (SVM) and random forests \cite[]{abraham2012sparsnp,abraham2014accurate,botta2014exploiting,okser2014regularized,wei2009disease}.

We recently developed two R packages, bigstatsr and bigsnpr, for efficiently analyzing large-scale genome-wide data \cite[]{prive2017efficient}. Package bigstatsr includes an efficient algorithm with a new implementation for computing sparse linear and logistic regressions on huge datasets as large as the UK Biobank \cite[]{bycroft2017genome}.
In this paper, we present a comprehensive comparative study of our implementation of penalized logistic regression against the C+T method and the T-Trees algorithm, a derivation of random forests that has shown high predictive performance \cite[]{botta2014exploiting}.
Note that in this comparison we do not include any LMM method for the reasons mentioned before and do not include any SVM method because it is expected to give similar results to logistic regression \cite[]{abraham2012sparsnp}. For the C+T model, we compare different thresholds of inclusion of SNPs. For the penalized logistic regression, we also introduce two improvements. First, we present a procedure that we call Cross-Model Selection and Averaging (CMSA) for automatically choosing the regularization parameter of the penalized logistic regression, which directly affects the number of SNPs included in the model. Secondly, we also show how to use feature engineering in order to capture not only linear effects, but also recessive and dominant effects.

For the simulations of this paper, we use real genotype data and simulate new phenotypes \cite{zhou2013polygenic}. 
In order to make our comparison as comprehensive as possible, we compare different disease architectures by varying the number, size and location of causal effects as well as the heritability. We also compare different models for simulating new phenotypes, one with only linear effects, and one which combines linear, dominant and interaction-type effects.
Overall, we found that the penalized logistic regression consistently achieves higher predictive performance while being faster than the C+T and T-Trees methods. This demonstrates the feasibility and relevance of this approach in PRS modeling in future studies.

@@

We use real genotypes of European individuals from a case-control celiac disease study \cite[]{dubois2010multiple}. The composition of this dataset is presented in table \ref{tab:celiac-data}. Details of quality control and imputation for this dataset are available in \cite{prive2017efficient}. For simulations presented later, we first restrict the dataset to controls in order to remove the genetic structure induced by the celiac disease status. Then, we decided to remove population structure because it can affect the predictive performance of methods \cite[]{martin2017human}.In order to alleviate population structure, we keep people from the UK only and we further remove outliers based on the robust Mahalanobis distances computed using the first 10 principal components of remaining individuals. This 3-step filtering process results in a sample of 7102 individuals with minimal population structure (see supplementary notebook ``preprocessing''). We also use this dataset for a real data application, in this case keeping all individuals.

@@

We simulate binary phenotypes using a Liability Threshold Model (LTM) with a prevalence of 30\% \cite[]{falconer1965inheritance}. We vary parameters of the simulations in order to match a range of genetic architecture, from low to high polygenicity. For practical reason, this was achieved by varying the number of chromosomes used (either all 22 autosomal chromosomes or chromosome 6 only), the number of causal variants and their location (30, 300 or 3000 anywhere on the genome or 30 only in the HLA region), and the heritability (50\% or 80\%). We also considered deviation form the standard normal additive model, drawing effects of causal SNPs from either Normal or Laplace distribution, and computing liability scores either from a ``simple'' model with linear effects only or a ``fancy'' model which combines linear, dominant and interaction-type effects. For the ``simple'' model, we compute $$y_i = \sum_{j\in S_\text{causal}} w_j \cdot \widetilde{G_{i,j}} ~~+~~ \epsilon_i$$ where $w_j$ are weights generated with a Gaussian or a Laplace distribution, $G_{i,j}$ is the allele count of individual $i$ for SNP $j$, $\widetilde{G_{i,j}}$ corresponds to its standardized version (zero mean and unit variance for all SNPs) and $\epsilon \sim N(0, 1 - h^2)$.
For the ``fancy'' model, we separate the causal SNPs in three equal sets $S_\text{causal}^{(1)}$, $S_\text{causal}^{(2)}$
and $S_\text{causal}^{(3)}$; $S_\text{causal}^{(3)}$ is further separated in two equal sets, $S_\text{causal}^{(3.1)}$ and $S_\text{causal}^{(3.2)}$.
We then compute $$y_i = \underbrace{\sum_{j\in S_\text{causal}^{(1)}} w_j \cdot \widetilde{G_{i,j}}}_\text{linear} + \underbrace{\sum_{j\in S_\text{causal}^{(2)}} w_j \cdot \widetilde{D_{i,j}}}_\text{dominant} + \underbrace{\sum_{\substack{k=1 \\ j_1=e_k^{(3.1)} \\ j_2=e_k^{(3.2)}}}^{k=\left|S_\text{causal}^{(3.1)}\right|} w_{j_1} \cdot \widetilde{G_{i,j_1} G_{i,j_2}}}_\text{interaction} ~~+~~ \epsilon_i$$
where $D_{i,j} = \mathds{1}\left\{G_{i,j} \neq 0\right\}$ and $S_\text{causal}^{(q)}=\left\{e_k^{(q)},~k \in \left\{1, \ldots, \left|S_\text{causal}^{(q)}\right|\right\}\right\}$.
Note that, because we used the raw genotype counts when coding the interaction terms, the corresponding SNPs also display a marginal effect.

We implement 3 different simulation scenarios, summarized in table \ref{tab:simus}.
Scenario \textnumero1 uses the whole dataset (all 22 autosomal chromosomes) and a training set of size 6000. It compares all methods described in section \ref{sec:methods}; for each combination of the remaining parameters, results are based on 100 simulations except  for T-Trees, which relies on 5 simulations only (and an heritability of 80\%) because of a much higher computational burden (several hours of computation for a single simulation) as compared to other approaches. Scenario \textnumero2 consists of 100 simulations per combination of parameters on a dataset composed of chromosome 6 only. Reducing the number of SNPs aims at increasing the polygenicity (the proportion of causal SNPs) of the simulated models and at virtually increasing the sample size \cite[]{dudbridge2013power,marquez2017multiethnic,vilhjalmsson2015modeling}.
For this scenario, we use only the additive (``simple'') model, but continue to compare all previous different values of the other parameters. Finally, scenario \textnumero3 reuses the whole dataset but this time varying the size of the training set in order to assess how the sample size affects predictive performance of methods. A total of 100 simulations per combination of parameters are run using 300 causal SNPs anywhere on the genome.

@@

In this study, we use two different measures of predictive accuracy. First, we use the Area Under the Receiver Operating Characteristic (ROC) Curve (AUC) \cite[]{fawcett2006introduction,lusted1971signal}. In the case of our study, the AUC is the probability that a PRS of a case is greater than the PRS of a control.
This measure indicates the extent to which we can distinguish between cases and controls using PRSs. As a second measure, the partial AUC for specificities between 90\% and 100\% is also reported \cite[]{dodd2003partial,mcclish1989analyzing}. This measure is similar to the AUC, but focuses on high specificities, which is the most useful part of the ROC curve in clinical settings. When reporting results of AUC, we use estimates of maximum achievable AUC values for the simulations: 84\% and 94\% for heritabilities of respectively 50\% and 80\% based on three different estimations (see Supplementary Materials). %Note that we also report the timing of the main computations and the number of SNPs used in the predictions.

@@

The C+T (Clumping + Thresholding) method directly derives a Polygenic Risk Score (PRS) from the results of a Genome-Wide Associations Study (GWAS). In the GWAS, a coefficient of regression (i.e.\ the estimated effect size $\hat\beta_j$) is learned independently for each SNP $j$ along with a corresponding p-value $p_j$. The SNPs are first clumped (C) so that there remain only loci that are weakly correlated with one another, which form a set of SNPs denoted $S_\text{clumping}$. Then, thresholding (T) consists in removing SNPs with p-values larger than a threshold $p_T$ to be determined. Finally, a PRS is defined as the sum of allele counts of the remaining SNPs weighted by the corresponding effect coefficients: $$\rm{PRS}_i = \sum_{\substack{j \in S_\text{clumping} \\ p_j~<~p_T}} \hat\beta_j \cdot G_{i,j}$$ where $\hat\beta_j$ ($p_j$) are the effect sizes (p-values) learned from the GWAS. In this study, we report scores for a clumping threshold at $r^2 > 0.2$ within regions of 500kb.

@@

Finally, for the penalized logistic regression, we solve: $$\arg\!\min_{\beta_0,~\beta}(\lambda, \alpha)\left\{  \underbrace{ -\sum_{i=1}^n \left( y_i \log\left(p_i\right) + (1 - y_i) \log\left(1 - p_i\right) \right) }_\text{Loss function}   +   \underbrace{ \lambda \left((1-\alpha)\frac{1}{2}\|\beta\|_2^2 + \alpha \|\beta\|_1\right) }_\text{Penalization}  \right\}$$ where $p_i=1/\left(1+\exp\left(-(\beta_0 + x_i^T\beta)\right)\right)$, $x$ is denoting the genotypes and covariables (e.g.\ principal components), $y$ is the disease status we want to predict, $\lambda$ and $alpha$ are two regularization parameters that need to be chosen.
Different regularizations can be used to prevent overfitting, among other benefits: the L2-regularization (``ridge'', \cite{hoerl1970ridge}) shrinks coefficients and is ideal if there are many predictors drawn from a Gaussian distribution (corresponds to $\alpha = 0$ in the previous equation);
the L1-regularization (``lasso'', \cite{tibshirani1996regression}) forces some of the coefficients to be equal to zero and can be used as a means of variable selection, leading to sparse models (corresponds to $\alpha = 1$);
the L1- and L2-regularization (``elastic-net'', \cite{zou2005regularization}) is a compromise between the two previous penalties and is particularly useful in the $m \gg n$ situation ($m$: number of SNPs), or any situation involving many correlated predictors (corresponds to $0 < \alpha < 1$) \cite[]{friedman2010regularization}. In this study, we use the elastic-net regularization with $\alpha = 0.5$.%, without trying to tune (e.g.\ by grid-search and cross-validation) the value of this hyper-parameter $\alpha$.

To fit this penalized logistic regression, we use a very efficient algorithm \cite[]{friedman2010regularization,tibshirani2012strong,zeng2017efficient} from which we derived our own implementation in the R package bigstatsr \cite[]{prive2017efficient}.
This type of algorithm builds predictions for many values of $\lambda$, which is called a ``regularization path''. To obtain an algorithm free of the choice of this hyper-parameter $\lambda$, we developed a procedure that we call Cross-Model Selection and Averaging (CMSA, figure \ref{fig:CMSA}).
Because of the L1-regularization, the resulting vectors of coefficients are typically very sparse and can be used to make a PRS based on a \textit{linear} combination of allele counts. We refer to this method as ``logit-simple'' in the results section.

In order to capture recessive and dominant effects in addition to linear effects, we use feature engineering: we construct a separate dataset with, for each SNP variable, two more variables coding for recessive and dominant effects. This results in a dataset with 3 times as many variables as the initial one, on which we can apply the penalized logistic regression with the CMSA procedure (Figure \ref{fig:CMSA}). We call this method ``logit-triple'' in the results.

@@

We use Monte Carlo cross-validation to compute AUC, partial AUC, number of predictors used and execution time for the original Celiac dataset with real phenotypes: we randomly split 100 times the dataset in a training set of 12,000 indiduals and a test set composed of the remaining 3155 individuals.

@@

We compared the penalized logistic regression (``logit-simple'') with the C+T method (``PRS'') using whole-genome simulations of scenario \textnumero1 (Table \ref{tab:simus}).

When simulating a model with 30 causal SNPs and an heritability of 80\%, penalized logistic regression provides AUC greater than 93\%, nearly reaching the maximum achievable AUC of 94\%, whereas AUC values obtained with C+T method range between 83\% and 90\% (Figure \ref{fig:main-AUC-logit}). 
Moreover, the penalized logistic regression consistently provides better predictive performance than the C+T method across all scenarios we considered, except in the case of high polygenicity and small sample size, in which case all methods hardly exceed 60\% of AUC (Figures \ref{fig:main-AUC-logit} and \ref{fig:supp-AUC-logit}).

Method ``logit-simple'' provides particularly higher predictive performance than ``PRS-max'' where there are correlations between predictors, i.e.\ when we choose causal SNPs to be in the HLA region. In this situation, the mean AUC reaches 92.5\% with the ``logit-simple'' and 83\% with the ``PRS-max'', for a maximum achievable AUC of 94\% (Figure \ref{fig:main-AUC-logit}).

@@

Method ``logit-simple'', that automatically chooses hyper-parameter $\lambda$, provides similar predictive performance than the best predictive performance of the biglasso implementation \cite[]{zeng2017efficient}, only slightly better for biglasso but likely due to over-fitting when reporting the best prediction (Figure \ref{fig:supp-biglasso}).

@@

We tested the T-Trees method in scenario \textnumero1.
As compared to ``logit-simple'', T-Trees perform worse in terms of predictive ability, while taking much longer to run and making more complex predictive models because T-Trees use more predictors and non-linear effects (Figure \ref{fig:supp-ttrees}). Even when simulating a fancy model in which there are dominant and interaction-type effects that T-Trees should be able to handle, AUC is still lower with T-Trees as compared with ``logit-simple''.

Secondly, we compare the two penalized logistic regressions in scenario \textnumero1, ``logit-simple'' and ``logit-triple'' that uses additional features (variables) coding for recessive and dominant effects.
Predictive performance of ``logit-triple'' are nearly as good as ``logit-simple'' when there are only linear effects (differences of AUC are always smaller than 2\%) and can lead to significantly greater results when there are also dominant and interactions effects (Figures \ref{fig:supp-triple} and \ref{fig:supp-AUC-triple}). For the ``fancy model'', ``logit-triple'' provides AUC values at least 3.5\% higher than ``logit-simple'', except when there are 3000 causal SNPs.
Yet, the ``triple'' solution takes 2-3 times as much time to run and requires 3 times as much disk storage as the ``logit-simple''.

Thirdly, for the ``fancy'' simulation model, we report similar results as in the previous subsection. As before, ``logit-simple'' achieves greater AUCs than the ``PRS-max'', except in the case of 3000 causal SNP anywhere on the genome, where no method exceed 60\% of AUC (Figure \ref{fig:supp-AUC-logit-fancy}). Moreover, the choice of the threshold inclusion of SNPs also greatly impacts the performance of the C+T method (Figure \ref{fig:supp-AUC-PRS-fancy}).

@@

First, when reproducing simulations of scenario \textnumero1 using only chromosome 6 (scenario \textnumero2), the predictive performance of all methods globally increase (Figure \ref{fig:supp-AUC-chr6}). 
There is particularly a large increase when simulating 3000 causal SNPs anywhere in the dataset: AUC from the ``logit-simple'' increases from less than 60\% to nearly 80\%. 
On the contrary, when simulating only 30 causal SNPs on the corresponding dataset, AUC of the ``PRS-max'' does not increase. 
Secondly, when varying the training size (scenario \textnumero3), we report an increase of AUC when increasing the training size, with a faster increase of AUC provided by ``logit-simple'' as compared to ``PRS-max'' (Figure  \ref{fig:main-AUC-ntrain}).

@@

In this comparative study, we present a computationally efficient implementation of penalized logistic regression. This model can be used to build polygenic risk scores based on very large SNP datasets such as the UK biobank \cite[]{bycroft2017genome}. In agreement with previous work \cite[]{abraham2013performance}, we show that jointly estimating SNP effects has the potential to dramatically improve predictive performance as compared to the standard C+T approach in which these effects are learned independently. Penalized logistic regression consistently outperform the C+T method, except in the case where there are only small effects and sample size is limited, in which case all methods perform poorly. Note that in this particular case, penalized logistic regression could outperform the C+T method if choosing a different value for hyper-parameter $\alpha$.

>>>>

In this paper, we present an efficient method to jointly estimate SNP effects, allowing for practical application of penalized logistic regression on modern datasets including hundreds of thousands of individuals. Our implementation of penalized logistic regression also includes nearly-optimal choices for hyper-parameters. The choice of hyper-parameters for a method is very important as it can dramatically impact the predictive performance of the method. As an example, AUC values range from less than 60\% to 90\% in a model with 30 causal SNPs, depending on the p-value threshold in C+T.

We compare the performance of penalized logistic regression to the C+T method and to a derivation of random forests. 
Penalized logistic regression consistently achieves higher predictive performance than the two other methods while being faster. We also found that improvement in predictive performance is more pronounced when there are few effects located in nearby genomic regions with correlated SNPs; AUC values increase from 83\% with C+T to 92.5\% with penalized logistic regression. We confirmed these results in a data analysis of a case-control study for celiac disease where penalized logistic regression and the standard C+T method achieve AUC of 89\% and of 82\%.

@@

Many methods have been developed to predict disease status, or more generally any phenotype, based on SNP information. A commonly used method, called ``P+T'' or ``C+T'' (which stands for ``Clumping and Thresholding'') -- or even just PRS -- is used to derive PRSs from results of Genome-Wide Association Studies (GWASs) \cite[]{chatterjee2013projecting,dudbridge2013power,evans2009harnessing,purcell2009common,wray2007prediction}. This technique uses GWAS summary statistics only, allowing for fast implementation. However, the ``C+T'' approach also has several limitations. Previous studies showed that predictive performance of the C+T method is very sensitive to the threshold of inclusion of SNPs, depending on the disease architecture \cite[]{ware2017heterogeneity}.
Linear Mixed-Models (LMMs) are another widely-used method in fields such as plant and animal breeding or for predicting highly heritable quantitative human phenotypes such as height \cite[]{lello2017accurate,yang2010common}. Yet, models resulting from LMM, known e.g.\ as ``gBLUP'', are not optimal for predicting disease status based on genotypes \cite[]{abraham2013performance}. Moreover, these methods and their derivatives are often computationally demanding, both in terms of memory and time required, which makes them unlikely to be used for prediction on very large datasets \cite[]{golan2014effective,maier2015joint,speed2014multiblup,zhou2013polygenic}.
Finally, statistical learning methods have also been used to derive PRSs for complex human diseases by jointly estimating SNP effects. Such methods include joint logistic regression, Support Vector Machine (SVM) and random forests \cite[]{abraham2012sparsnp,abraham2014accurate,botta2014exploiting,okser2014regularized,wei2009disease}.

We recently developed two R packages, bigstatsr and bigsnpr, for efficiently analyzing large-scale genome-wide data \cite[]{prive2017efficient}. Package bigstatsr includes an efficient algorithm with a new implementation for computing sparse linear and logistic regressions on huge datasets as large as the UK Biobank \cite[]{bycroft2017genome}.
In this paper, we present a comprehensive comparative study of our implementation of penalized logistic regression against the C+T method and the T-Trees algorithm, a derivation of random forests that has shown high predictive performance \cite[]{botta2014exploiting}.
In this comparison, we do not include any LMM method for the reasons mentioned before and do not include any SVM method because it is expected to give similar results to logistic regression \cite[]{abraham2012sparsnp}. For the C+T model, we compare different thresholds of inclusion of SNPs. For the penalized logistic regression, we also introduce two improvements. First, we present a procedure that we call Cross-Model Selection and Averaging (CMSA) for automatically choosing the regularization parameter of the penalized logistic regression, which directly affects the number of SNPs included in the model. Secondly, we show how to use feature engineering in order to capture not only linear effects, but also recessive and dominant effects.

For the simulations of this paper, we use real genotype data and simulate new phenotypes \cite[]{zhou2013polygenic}. 
In order to make our comparison as comprehensive as possible, we compare different disease architectures by varying the number, size and location of causal effects as well as the heritability. We also compare different models for simulating new phenotypes, one with only linear effects, and one which combines linear, dominant and interaction-type effects.
Overall, we found that the penalized logistic regression consistently achieves higher predictive performance while being faster than the C+T and T-Trees methods. This demonstrates the feasibility and relevance of this approach for PRS computation.

@@

We use real genotypes of European individuals from a case-control celiac disease study \cite[]{dubois2010multiple}. The composition of this dataset is presented in table \ref{tab:celiac-data}. Details of quality control and imputation for this dataset are available in \cite{prive2017efficient}. For simulations presented later, we first restrict the dataset to controls in order to remove the genetic structure induced by the celiac disease status. Then, we decided to remove population structure because it can affect the predictive performance of methods \cite[]{martin2017human}.
In order to alleviate population structure, we keep people from the UK only and we further remove outliers based on the robust Mahalanobis distances computed using the first 10 principal components of the remaining individuals. This 3-step filtering process results in a sample of 7102 individuals with minimal population structure (see supplementary notebook ``preprocessing''). We also use this dataset for real data application, in this case keeping all 15,155 individuals.

@@

We simulate binary phenotypes using a Liability Threshold Model (LTM) with a prevalence of 30\% \cite[]{falconer1965inheritance}. We vary parameters of the simulations in order to match a range of genetic architecture, from low to high polygenicity. 
This was achieved by varying the number of causal variants and their location (30, 300, or 3000 anywhere or 30 in the HLA region only), and the heritability (50\% or 80\%). 
Then, to further increase the proportion of causal variants, instead of using all 22 automosal chromosomes, we restricted the dataset to chromosome 6 only.
We also considered deviation form the standard normal additive model, drawing effects of causal SNPs from either a Normal or from a Laplace distribution, and computing liability scores either from a ``simple'' model with linear effects only or a ``fancy'' model which combines linear, dominant and interaction-type effects. For the ``simple'' model, we compute the liability score of the i-th individual $$y_i = \sum_{j\in S_\text{causal}} w_j \cdot \widetilde{G_{i,j}} ~~+~~ \epsilon_i~,$$ where $w_j$ are weights generated with a Gaussian or a Laplace distribution, $G_{i,j}$ is the allele count of individual $i$ for SNP $j$, $\widetilde{G_{i,j}}$ corresponds to its standardized version (zero mean and unit variance for all SNPs) and $\epsilon \sim N(0, 1 - h^2)$.
For the ``fancy'' model, we separate the causal SNPs in three equal sets $S_\text{causal}^{(1)}$, $S_\text{causal}^{(2)}$
and $S_\text{causal}^{(3)}$; $S_\text{causal}^{(3)}$ is further separated in two equal sets, $S_\text{causal}^{(3.1)}$ and $S_\text{causal}^{(3.2)}$.
We then compute $$y_i = \underbrace{\sum_{j\in S_\text{causal}^{(1)}} w_j \cdot \widetilde{G_{i,j}}}_\text{linear} + \underbrace{\sum_{j\in S_\text{causal}^{(2)}} w_j \cdot \widetilde{D_{i,j}}}_\text{dominant} + \underbrace{\sum_{\substack{k=1 \\ j_1=e_k^{(3.1)} \\ j_2=e_k^{(3.2)}}}^{k=\left|S_\text{causal}^{(3.1)}\right|} w_{j_1} \cdot \widetilde{G_{i,j_1} G_{i,j_2}}}_\text{interaction} ~~+~~ \epsilon_i~,$$
where $D_{i,j} = \mathds{1}\left\{G_{i,j} \neq 0\right\}$ and $S_\text{causal}^{(q)}=\left\{e_k^{(q)},~k \in \left\{1, \ldots, \left|S_\text{causal}^{(q)}\right|\right\}\right\}$.
Note that, because we used the raw genotype counts when coding the interaction terms, the corresponding SNPs also display a marginal effect.

We implement 3 different simulation scenarios, summarized in table \ref{tab:simus}.
Scenario \textnumero1 uses the whole dataset (all 22 autosomal chromosomes) and a training set of size 6000. It compares all methods described in section \ref{sec:methods}; for each combination of the remaining parameters, results are based on 100 simulations except  for T-Trees, which relies on 5 simulations only (and an heritability of 80\%) because of a much higher computational burden (several hours of computation for a single simulation) as compared to other approaches. Scenario \textnumero2 consists of 100 simulations per combination of parameters on a dataset composed of chromosome 6 only. Reducing the number of SNPs aims at increasing the polygenicity (the proportion of causal SNPs) of the simulated models and at virtually increasing the sample size \cite[]{dudbridge2013power,marquez2017multiethnic,vilhjalmsson2015modeling}.
For this scenario, we use only the additive (``simple'') model, but continue to compare all previous different values of the other parameters. Finally, scenario \textnumero3 reuses the whole dataset but this time varying the size of the training set in order to assess how the sample size affects predictive performance of methods. A total of 100 simulations per combination of parameters are run using 300 causal SNPs randomly chosen anywhere on the genome.

@@

In this study, we use two different measures of predictive accuracy. First, we use the Area Under the Receiver Operating Characteristic (ROC) Curve (AUC) \cite[]{fawcett2006introduction,lusted1971signal}. In the case of our study, the AUC is the probability that a PRS of a case is greater than the PRS of a control.
This measure indicates the extent to which we can distinguish between cases and controls using PRSs. As a second measure, the partial AUC for specificities between 90\% and 100\% is also reported \cite[]{dodd2003partial,mcclish1989analyzing}. This measure is similar to the AUC, but focuses on high specificities, which is the most useful part of the ROC curve in clinical settings. 
When reporting AUC results of simulations, we use estimates of maximum achievable AUC of 84\% and 94\% for heritabilities of respectively 50\% and 80\%, based on three different yet consistent estimations (see Supplementary Materials). %Note that we also report the timing of the main computations and the number of SNPs used in the predictions.

@@

The C+T (Clumping + Thresholding) method directly derives a Polygenic Risk Score (PRS) from the results of a Genome-Wide Associations Study (GWAS). In GWAS, a coefficient of regression (i.e.\ the estimated effect size $\hat\beta_j$) is learned independently for each SNP $j$ along with a corresponding p-value $p_j$. The SNPs are first clumped (C) so that there remain only loci that are weakly correlated with one another, which form a set of SNPs denoted $S_\text{clumping}$. Then, thresholding (T) consists in removing SNPs with p-values larger than a threshold $p_T$ to be determined. Finally, a PRS is defined as the sum of allele counts of the remaining SNPs weighted by the corresponding effect coefficients: $$\rm{PRS}_i = \sum_{\substack{j \in S_\text{clumping} \\ p_j~<~p_T}} \hat\beta_j \cdot G_{i,j}~,$$ where $\hat\beta_j$ ($p_j$) are the effect sizes (p-values) learned from the GWAS. In this study, we report scores for a clumping threshold at $r^2 > 0.2$ within regions of 500kb.

@@

Finally, for the penalized logistic regression, we find regression coefficients $\beta_0$ and $\beta$ that minimize the following regularized loss function: $$L(\lambda, \alpha) = \underbrace{ -\sum_{i=1}^n \left( y_i \log\left(p_i\right) + (1 - y_i) \log\left(1 - p_i\right) \right) }_\text{Loss function}   +   \underbrace{ \lambda \left((1-\alpha)\frac{1}{2}\|\beta\|_2^2 + \alpha \|\beta\|_1\right) }_\text{Penalization} ~,$$ where $p_i=1/\left(1+\exp\left(-(\beta_0 + x_i^T\beta)\right)\right)$, $x$ is denoting the genotypes and covariables (e.g.\ principal components), $y$ is the disease status we want to predict, $\lambda$ and $\alpha$ are two regularization parameters that need to be chosen.
Different regularizations can be used to prevent overfitting, among other benefits: the L2-regularization (``ridge'', \cite{hoerl1970ridge}) shrinks coefficients and is ideal if there are many predictors drawn from a Gaussian distribution (corresponds to $\alpha = 0$ in the previous equation);
the L1-regularization (``lasso'', \cite{tibshirani1996regression}) forces some of the coefficients to be equal to zero and can be used as a means of variable selection, leading to sparse models (corresponds to $\alpha = 1$);
the L1- and L2-regularization (``elastic-net'', \cite{zou2005regularization}) is a compromise between the two previous penalties and is particularly useful in the $m \gg n$ situation ($m$: number of SNPs), or any situation involving many correlated predictors (corresponds to $0 < \alpha < 1$) \cite[]{friedman2010regularization}. In this study, we use elastic-net regularization with $\alpha = 0.5$.%, without trying to tune (e.g.\ by grid-search and cross-validation) the value of this hyper-parameter $\alpha$.

To fit this penalized logistic regression, we use a very efficient algorithm \cite[]{friedman2010regularization,tibshirani2012strong,zeng2017efficient} from which we derived our own implementation in the R package bigstatsr \cite[]{prive2017efficient}.
This type of algorithm builds predictions for many values of $\lambda$, which is called a ``regularization path''. To obtain an algorithm free of the choice of this hyper-parameter $\lambda$, we developed a procedure that we call Cross-Model Selection and Averaging (CMSA, figure \ref{fig:CMSA}).
Because of L1-regularization, the resulting vectors of coefficients are typically very sparse and can be used to make a PRS based on a \textit{linear} combination of allele counts. We refer to this method as ``logit-simple'' in the results section.

In order to capture recessive and dominant effects in addition to linear effects, we use feature engineering: we construct a separate dataset with, for each SNP variable, two more variables coding for recessive and dominant effects. This results in a dataset with 3 times as many variables as the initial one, on which we can apply penalized logistic regression with the CMSA procedure (Figure \ref{fig:CMSA}). We call this method ``logit-triple'' in the results.

@@

We use Monte Carlo cross-validation to compute AUC, partial AUC, the number of predictors and execution time for the original Celiac dataset with real phenotypes: we randomly split 100 times the dataset in a training set of 12,000 indiduals and a test set composed of the remaining 3155 individuals.

@@

We compared penalized logistic regression (``logit-simple'') with the C+T method (``PRS'') using whole-genome simulations of scenario \textnumero1 (Table \ref{tab:simus}).

When simulating a model with 30 causal SNPs and an heritability of 80\%, penalized logistic regression provides AUC greater than 93\%, nearly reaching the maximum achievable AUC of 94\%, whereas AUC values obtained with C+T method range between 83\% and 90\% (Figure \ref{fig:main-AUC-logit}). 
Moreover, penalized logistic regression consistently provides better predictive performance than the C+T method across all scenarios we considered, except in the case of high polygenicity and small sample size, in which case all methods hardly exceed 60\% of AUC (Figures \ref{fig:main-AUC-logit} and \ref{fig:supp-AUC-logit}).

Method ``logit-simple'' provides particularly higher predictive performance than ``PRS-max'' where there are correlations between predictors, i.e.\ when we choose causal SNPs to be in the HLA region. In this situation, the mean AUC reaches 92.5\% with the ``logit-simple'' approach and 83\% with ``PRS-max'', and the maximum achievable AUC is of 94\% (Figure \ref{fig:main-AUC-logit}).

@@

Method ``logit-simple'' that automatically chooses hyper-parameter $\lambda$ provides similar predictive performance than the best predictive performance of the biglasso implementation \cite[]{zeng2017efficient}, only slightly better for biglasso but likely due to over-fitting when reporting the best prediction (Figure \ref{fig:supp-biglasso}).

@@

We tested the T-Trees method in scenario \textnumero1.
As compared to ``logit-simple'', T-Trees perform worse in terms of predictive ability, while taking much longer to run and making more complex predictive models because T-Trees use more predictors and non-linear effects (Figure \ref{fig:supp-ttrees}). Even when simulating a fancy model in which there are dominant and interaction-type effects that T-Trees should be able to handle, AUC is still lower with T-Trees as compared with ``logit-simple'' (Figure \ref{fig:supp-ttrees}).

Secondly, we compare the two penalized logistic regressions in scenario \textnumero1, ``logit-simple'' and ``logit-triple'' that uses additional features (variables) coding for recessive and dominant effects.
Predictive performance of ``logit-triple'' are nearly as good as ``logit-simple'' when there are only linear effects (differences of AUC are always smaller than 2\%) and can lead to significantly greater results when there are also dominant and interactions effects (Figures \ref{fig:supp-AUC-triple} and \ref{fig:supp-triple}). For the ``fancy model'', ``logit-triple'' provides AUC values at least 3.5\% higher than ``logit-simple'', except when there are 3000 causal SNPs.
Yet, the ``triple'' solution takes 2-3 times as much time to run and requires 3 times as much disk storage as the ``logit-simple''.

Thirdly, for the ``fancy'' simulation model, we report similar results as in the previous subsection. As before, ``logit-simple'' achieves greater AUCs than the ``PRS-max'', except in the case of 3000 causal SNPs randomly chosen anywhere on the genome, where no method exceed 60\% of AUC (Figure \ref{fig:supp-AUC-logit-fancy}). Moreover, the choice of the threshold inclusion of SNPs also greatly impacts the performance of the C+T method (Figure \ref{fig:supp-AUC-PRS-fancy}).

@@

First, when reproducing simulations of scenario \textnumero1 using only chromosome 6 (scenario \textnumero2), the predictive performance of all methods globally increase (Figure \ref{fig:supp-AUC-chr6}). 
There is particularly a large increase when simulating 3000 causal SNPs: AUC from the ``logit-simple'' increases from less than 60\% to nearly 80\%. 
On the contrary, when simulating only 30 causal SNPs on the corresponding dataset, AUC of the ``PRS-max'' does not increase. 
Secondly, when varying the training size (scenario \textnumero3), we report an increase of AUC when increasing the training size, with a faster increase of AUC provided by ``logit-simple'' as compared to ``PRS-max'' (Figure  \ref{fig:main-AUC-ntrain}).

@@

In this comparative study, we present a computationally efficient implementation of penalized logistic regression. 
%This model can be used to build polygenic risk scores based on very large SNP datasets such as the UK biobank \cite[]{bycroft2017genome}. 
In agreement with previous work \cite[]{abraham2013performance}, we show that jointly estimating SNP effects has the potential to substantially improve predictive performance as compared to the standard C+T approach in which SNP effects are learned independently. Penalized logistic regression consistently outperform the C+T method, except in the case where there are only small effects and sample size is limited, in which case all methods perform poorly. Note that in this particular case, penalized logistic regression could outperform the C+T method if choosing $\alpha = 10^{-3}$, which gives more importance to ridge regularization and accept more predictors in the model.
